---
title: "Body Performance Classification"
output:
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r include=FALSE}
library(tidyverse)
library(janitor)       
library(corrplot)       
library(gridExtra)     
library(nnet)          
library(caret)        
library(rpart.plot)    
library(randomForest)
library(xgboost)
library(themis)
library(tidymodels)
library(VIM)
library(lmPerm)
```

## Read data

```{r}
data <- read.csv("bodyPerformance.csv")
data <- clean_names(data)
```

------------------------------------------------------------------------

## **Statistical description of the dataset**

### Check the structure of the data

```{r}
dim(data)
glimpse(data)
```

### Statistical description of the data

```{r}
summary(data)
```

### Define numerical and categorical features

```{r}
numerical_features <- c("age", "height_cm", "weight_kg", "body_fat", "diastolic", "systolic",
                        "grip_force", "sit_and_bend_forward_cm", "sit_ups_counts",
                        "broad_jump_cm")

categorical_features <- c("gender", "class")
```

------------------------------------------------------------------------

## **Data preprocessing**

### Check missing data

```{r}
na_percentage <- colSums(is.na(data)) / nrow(data) * 100
na_percentage

aggr(data, ylab = c("Proportion of missings", "Pattern"), number = TRUE,
cex.axis = 0.6, cex.numbers = 0.5)
```

-   There is no missing data.

### Check duplicate data

```{r}
sum(duplicated(data))
data[duplicated(data), ]
```

-   There is 1 duplicate row.
-   With only 1 duplicate row, the impact on the overall analysis may not be significant. However, keeping duplicate rows may lead to slight bias in descriptive statistics or modeling.

#### Remove duplicate rows

```{r}
data <- data[!duplicated(data), ]

dim(data)

sum(duplicated(data))
```

### Check outliers

```{r, fig.width=15, fig.height=12}
boxplot_list <- list()

for (feature in numerical_features) {
  p <- ggplot(data, aes(x = "", y = .data[[feature]])) +
    geom_boxplot(fill = "lightblue", color = "darkblue", alpha = 0.7, 
                 width = 0.5, outlier.size = 3) +  
    labs(title = feature) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      axis.title = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_text(size = 12) 
    )
  boxplot_list[[feature]] <- p
}


grid.arrange(
  grobs = boxplot_list, 
  ncol = 3, 
  nrow = 4,
  top = textGrob(
    "Boxplots of Numerical Features", 
    gp = gpar(fontsize = 20, fontface = "bold")  
  ),
  padding = unit(2, "line")  
)
```

-   Quick note:
    -   Most variables have outliers
    -   Special variables diastolic, systolic, body_fat, sit_and_bend_forward_cm have outliers that are unreasonable with the real world
    -   Consider removing outliers or replacing them

#### Handle outliers

```{r}
# Remove outliers based on IQR
remove_outliers <- function(column) {
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  column >= lower_bound & column <= upper_bound
}

# Winsorization function
winsorize <- function(column) {
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  column[column < lower_bound] <- lower_bound
  column[column > upper_bound] <- upper_bound
  return(column)
}

# Start processing data
data <- data %>%
  # Remove outliers of diastolic and systolic
  filter(remove_outliers(diastolic), remove_outliers(systolic)) %>%
  # Remove negative values and outliers of sit_and_bend_forward_cm
  filter(sit_and_bend_forward_cm >= 0) %>%
  filter(remove_outliers(sit_and_bend_forward_cm)) %>%
  # BWinsorization cho body_fat
  mutate(body_fat = winsorize(body_fat))


summary(data)
```

```{r}
for (feature in c("diastolic", "systolic", "sit_and_bend_forward_cm", "body_fat")) {
  print(
    ggplot(data, aes(x = "", y = .data[[feature]])) +
      geom_boxplot(fill = "skyblue", color = "black") +
      labs(title = paste("Boxplot after Outlier Handling:", feature), x = "", y = feature) +
      theme_minimal()
  )
}
```

## **Visualize cleaned data**

### Gender and class distribution

```{r}
ggplot(data, aes(x = gender)) +
  geom_bar(fill = "blue") +
  coord_flip() +  
  geom_text(stat = "count", aes(label = after_stat(count)), hjust = -0.2, color = "white") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 15, face = "bold"))
```

```{r}
ggplot(data, aes(x = class)) +
  geom_bar(fill = "blue") +
  coord_flip() +  
  geom_text(stat = "count", aes(label = after_stat(count)), hjust = -0.2, color = "white") +
  labs(title = "Class Distribution", x = "Performance Class", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 15, face = "bold"))

```

```{r}
table( data$class)
```

-   After processing, class D is imbalanced compared to other classes.

### Frequency and density plot of each numerical variable

```{r, fig.width=15, fig.height=12}
hist_plots <- list()

for (var in numerical_features) {
  p <- ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(aes(y = after_stat(density)), 
                  fill = "skyblue", 
                  color = "darkblue",
                  alpha = 0.7,
                  bins = 30) +
    geom_density(color = "red", linewidth = 1.2) +  
    labs(title = var) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 12),
      axis.text = element_text(size = 12)
    )
  hist_plots[[var]] <- p
}

grid.arrange(
  grobs = hist_plots, 
  ncol = 3, 
  nrow = 4,
  top = textGrob(
    "Distribution of Numerical Features", 
    gp = gpar(fontsize = 20, fontface = "bold")
  ),
  padding = unit(2, "line")
)
```

### Correlation heatmap

```{r}
numeric_df <- data[, numerical_features]
cor_body <- cor(numeric_df, method = "pearson")
cor_body
```

```{r}
corrplot(cor_body)
```

## Perform some tests

### Boxplot and violin plot for variables

```{r, fig.width=15, fig.height=12}
violin_box_list <- list()

for(var in numerical_features) {
  violin_box_list[[var]] <- ggplot(data, aes(x = class, y = .data[[var]], fill = class)) +
    geom_violin(alpha = 0.5) +
    geom_boxplot(width = 0.2, alpha = 0.7) +
    scale_fill_brewer(palette = "Set3") +  
    labs(
      title = paste(var),
      x = "Performance Class",
      y = var
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, hjust = 0.5),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      legend.position = "none",
      panel.grid.major.y = element_line(color = "grey90", size = 0.3)
    )
}

grid.arrange(
  grobs = violin_box_list, 
  ncol = 3, 
  nrow = 4,
  top = textGrob(
    "Distribution by Performance Class",
    gp = gpar(fontsize = 16, fontface = "bold")
  ),
  padding = unit(2, "line")
)
```

### Compare variables between performance classes

```{r}
set.seed(54)

anova_results <- list()
plot_list <- list()

for(i in numerical_features) {
    out_aov <- aovp(formula = data[[i]] ~ class, data = data, perm = "Prob")
    anova_results[[i]] <- summary(out_aov)
    
    p <- ggplot(data, aes(x = class, y = .data[[i]], fill = class)) +
        geom_boxplot() +
        labs(title = paste("Distribution of", i, "by Performance Class"),
             x = "Performance Class",
             y = i) +
        theme_minimal() +
        theme(legend.position = "none") +
        scale_fill_brewer(palette = "Set3")
    
    plot_list[[i]] <- p
    
    # ANOVA results
    print(paste("ANOVA Results for", i))
    print(summary(out_aov))
    cat("\n--------------------------------------------------------------\n")
}

anova_summary <- data.frame(
    Variable = character(),
    F_value = numeric(),
    P_value = numeric(),
    Significant = character(),
    stringsAsFactors = FALSE
)

```

-   All variables have very small p-value, indicating that the difference is statistically significant.

-   **Analysis by variable group**:

    -   **Strength indicators** (grip_force, broad_jump_cm):

        -   Clear difference between classes

        -   Class A and B show outstanding superiority

-   **Endurance indicators** (sit_ups_counts):

    -   Clear difference between classes

    -   Class A shows outstanding superiority

-   **Body composition indicators** (body_fat, weight_kg):

    -   Clear difference between classes

    -   Class A shows outstanding superiority

**Meaning**: - All variables have very small p-value, indicating that the difference is statistically significant. - Strength indicators (grip_force, broad_jump_cm): Clear difference between classes - Class A and B show outstanding superiority - Endurance indicators (sit_ups_counts): Clear difference between classes - Class A shows outstanding superiority - Body composition indicators (body_fat, weight_kg): Clear difference between classes - Class A shows outstanding superiority

#### Analyze the characteristics of the high performance group (Class A)

```{r}
# Compare indices between Class A and other classes
class_comparison <- data %>%
  mutate(performance_group = ifelse(class == "A", "High Performance", "Others")) %>%
  group_by(performance_group) %>%
  summarise(
    avg_age = mean(age),
    avg_body_fat = mean(body_fat),
    avg_grip_force = mean(grip_force),
    avg_sit_ups = mean(sit_ups_counts),
    avg_broad_jump = mean(broad_jump_cm)
  )

# Visualize the difference
class_comparison_long <- class_comparison %>%
  pivot_longer(cols = starts_with("avg_"),
               names_to = "metric",
               values_to = "value")

ggplot(class_comparison_long, 
       aes(x = metric, y = value, fill = performance_group)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Characteristics of High Performance Group vs Others",
       x = "Metrics",
       y = "Average Value") +
  theme_minimal()
```

### Analyze the impact of gender on performance

```{r}
important_vars <- c("age", "body_fat", "grip_force", "sit_ups_counts", "broad_jump_cm")

# Perform permutation ANOVA for gender x class interaction
cat("Two-way permutation ANOVA (Gender x Class):\n")

for(var in important_vars) {
    formula <- as.formula(paste(var, "~ gender * class"))
    out_aov <- aovp(formula, data = data, perm = "Prob")
    cat(sprintf("\nTwo-way permutation ANOVA for %s:\n", var))
    print(summary(out_aov))
}

data_long <- data %>%
  pivot_longer(cols = all_of(important_vars),
               names_to = "variable",
               values_to = "value")

ggplot(data_long, aes(x = class, y = value, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Distribution of metrics by class and gender",
       x = "Class",
       y = "Value") +
  theme_minimal()

performance_stats <- data %>%
  group_by(gender, class) %>%
  summarise(
    count = n(),
    avg_age = mean(age),
    avg_body_fat = mean(body_fat),
    avg_grip_force = mean(grip_force),
    avg_sit_ups = mean(sit_ups_counts),
    avg_broad_jump = mean(broad_jump_cm)
  ) %>%
  group_by(gender) %>%
  mutate(percentage = count/sum(count) * 100)

print("\nStatistics by gender and class:")
print(performance_stats)

```

**Note:**

1.  **Role of gender in physical performance**:
    -   Gender is a strong factor affecting all physical indices.
    -   All p-values \< 2.2e-16, indicating that the difference between men and women is statistically significant.
    -   Men generally have an advantage in physical performance, particularly in indices such as grip strength and jump distance.
2.  **Class differentiation**:
    -   There is a clear difference between classes in all physical indices.
    -   The gap in physical performance between men and women increases as the class level increases.
    -   Class A stands out with outstanding physical indices and low body fat percentage.
3.  **Gender interaction with class**:
    -   The difference between men and women is not uniform across classes.
    -   The gap in physical performance between men and women increases as the class level increases.
    -   This suggests that men and women may need different training programs to optimize development.
4.  **Practical meaning**:
    -   Need to build individual training programs for men and women.

------------------------------------------------------------------------

## **Model**

Model evaluation function

```{r}
eval_multi_class <- function(x) {
  cc <- sum(diag(x))               
  sc <- sum(x)                      
  pp <- colSums(x)                 
  tt <- rowSums(x)                 
  
  # Precision and Recall for each class
  prec <- diag(x) / pp
  recall <- diag(x) / tt
  
  # Macro Precision, Recall and F1
  macro_prec <- mean(prec, na.rm = TRUE)
  macro_recall <- mean(recall, na.rm = TRUE)
  macro_f1 <- 2 * macro_prec * macro_recall / (macro_prec + macro_recall)
  
  # Accuracy
  acc <- cc / sc
  
  # Kappa
  kap <- (cc * sc - sum(pp * tt)) / (sc^2 - sum(pp * tt))
  
  return(list(
    Precision = prec,
    Recall = recall,
    Accuracy = acc,
    Kappa = kap,
    Macro_F1 = macro_f1
  ))
}

```

### Data split

```{r}
set.seed(123)
split <- initial_split(data, prop = 0.8, strata = class)
train <- training(split)
test <- testing(split)

cat("Train size: ", dim(train), "\n")
cat("Test size: ", dim(test), "\n")

```

```{r}
# Check data type
str(train$class)
str(test$class)

# Convert categorical variables to factor
train <- train %>%
  mutate(
    class = as.factor(class),
    gender = as.factor(gender)
  )

test <- test %>%
  mutate(
    class = as.factor(class),
    gender = as.factor(gender)
  )
str(train$class)
str(test$class)
```

### Balance data with SMOTE

```{r}
library(themis)
library(recipes)
library(tidymodels)

original_distribution <- table(train$class)

recipe_obj <- recipe(class ~ ., data = train) %>%
  step_smotenc(class)  

data_balanced <- prep(recipe_obj) %>%
  bake(new_data = NULL)

cat("Distribution of class before balancing:\n")
print(original_distribution)

cat("\nDistribution of class after balancing:\n")
print(table(data_balanced$class))

# Visualization
p1 <- ggplot(train, aes(x = class)) +
  geom_bar(fill = "skyblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Before balancing",
       x = "Class",
       y = "Count") +
  theme_minimal()

p2 <- ggplot(data_balanced, aes(x = class)) +
  geom_bar(fill = "salmon") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "After balancing",
       x = "Class",
       y = "Count") +
  theme_minimal()

# Display both plots
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)

summary(data_balanced)


data <- data_balanced

```

### Multinomial Logistic Regression

#### Multinomial Logistic Regression without feature transformation

```{r}
multinom_model <- multinom(class ~ ., data = train, maxit = 100)
pred <-  predict(multinom_model, newdata = test, type = "class")
head(pred)
```

```{r}
# confusion matrix
multinom_conf_mat <- tibble(
  truth = test$class,
  prediction = pred
) %>%
  conf_mat(truth = truth, estimate = prediction)

# Evaluate model
metrics_multinom <- eval_multi_class(multinom_conf_mat$table)
print(metrics_multinom)

# Visualization
autoplot(multinom_conf_mat, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix - Multinomial Logistic Regression")
```

#### Train Multinomial Logistic Regression model with interaction, log, and square transformations

```{r}
multinom_model_simplified <- multinom(
  class ~ gender + age + diastolic + systolic + height_cm +
    I(log(grip_force + 1)) +
    I(log(sit_ups_counts + 1)) +
    I(log(broad_jump_cm + 1)) +
    I(weight_kg * (body_fat/100)) +
    sit_and_bend_forward_cm,
  data = train, 
  maxit = 1000
)
pred_simplified <- predict(multinom_model_simplified, newdata = test, type = "class")
head(pred_simplified)
```

```{r}
# Confusion Matrix
multinom_conf_mat_simplified <- tibble(
  truth = test$class,
  prediction = pred_simplified
) %>%
  conf_mat(truth = truth, estimate = prediction)

print(multinom_conf_mat_simplified)
# Evaluate model
metrics_simplified <-eval_multi_class(multinom_conf_mat_simplified$table)
print(metrics_simplified)

# Visualization
autoplot(multinom_conf_mat_simplified, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix - Simplified Multinomial Model")

```

### Random Forest

```{r}
library(randomForest)
rf_model <- randomForest(class ~ ., data = train)
pred_rf <- predict(rf_model, newdata = test, type = "class")
head(pred_rf)
```

```{r}
# Confusion Matrix
rf_conf_mat <- tibble(
  truth = test$class,
  prediction = pred_rf
) %>%
  conf_mat(truth = truth, estimate = prediction)

print(rf_conf_mat)

# Evaluate model
metrics_rf <- eval_multi_class(rf_conf_mat$table)
print(metrics_rf)

# Visualization
autoplot(rf_conf_mat, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix - Random Forest")
```

### Training XGBoost

```{r}
library(xgboost)

# Prepare data
# Encoding class (A=0, B=1, C=2, D=3)
train$class_num <- as.numeric(factor(train$class, levels = c("A", "B", "C", "D"))) - 1
test$class_num <- as.numeric(factor(test$class, levels = c("A", "B", "C", "D"))) - 1

# Encoding gender (Female=0, Male=1)
train$gender_num <- ifelse(train$gender == "M", 1, 0)
test$gender_num <- ifelse(test$gender == "M", 1, 0)

# Choose numeric features
feature_cols <- c("age", "height_cm", "weight_kg", "body_fat", "diastolic", 
                 "systolic", "grip_force", "sit_and_bend_forward_cm", 
                 "sit_ups_counts", "broad_jump_cm", "gender_num")

# Convert data to matrix
train_matrix <- as.matrix(train[, feature_cols])
test_matrix <- as.matrix(test[, feature_cols])

# Train model
xgb_model <- xgboost(data = train_matrix,
                     label = train$class_num,
                     nrounds = 100,
                     objective = "multi:softmax",
                     num_class = 4,
                     eta = 0.3,
                     max_depth = 6)

# Predict
pred_xgb_num <- predict(xgb_model, test_matrix)
# Convert prediction to factor with correct order
pred_xgb <- factor(pred_xgb_num, levels = 0:3, labels = c("A", "B", "C", "D"))
```

```{r}
# Confusion Matrix
xgb_conf_mat <- tibble(
  truth = test$class,
  prediction = factor(pred_xgb, levels = levels(test$class))
) %>%
  conf_mat(truth = truth, estimate = prediction)

print(xgb_conf_mat)

# Evaluate model
metrics_xgb <- eval_multi_class(xgb_conf_mat$table)
print(metrics_xgb)

# Visualization
autoplot(xgb_conf_mat, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix - XGBoost")
```

### Model comparison

```{r}
comparison <- data.frame(
  Model = c("Basic Multinomial", "Simplified Multinomial", "Random Forest", "XGBoost"),
  Accuracy = c(metrics_multinom$Accuracy, metrics_simplified$Accuracy, 
               metrics_rf$Accuracy, metrics_xgb$Accuracy),
  Kappa = c(metrics_multinom$Kappa, metrics_simplified$Kappa, 
            metrics_rf$Kappa, metrics_xgb$Kappa),
  Macro_F1 = c(metrics_multinom$Macro_F1, metrics_simplified$Macro_F1, 
               metrics_rf$Macro_F1, metrics_xgb$Macro_F1)
)
print(comparison)
```

### Feature importance

```{r}
# 1.Multinomial
basic_importance <- varImp(multinom_model)
ggplot(data = data.frame(
  Feature = rownames(basic_importance),
  Importance = basic_importance[,1]
)) +
  geom_bar(aes(x = reorder(Feature, Importance), y = Importance), stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Feature Importance - Basic Multinomial",
       x = "Features", y = "Importance") +
  theme_minimal()

# 2. Simplified Multinomial
simplified_importance <- varImp(multinom_model_simplified)
ggplot(data = data.frame(
  Feature = rownames(simplified_importance),
  Importance = simplified_importance[,1]
)) +
  geom_bar(aes(x = reorder(Feature, Importance), y = Importance), stat = "identity", fill = "lightgreen") +
  coord_flip() +
  labs(title = "Feature Importance - Simplified Multinomial",
       x = "Features", y = "Importance") +
  theme_minimal()

# 3. Random Forest
varImpPlot(rf_model, 
           sort = TRUE,
           main = "Feature Importance - Random Forest",
           n.var = min(10, ncol(train)))

# 4. XGBoost
importance_matrix <- xgb.importance(
  feature_names = feature_cols,
  model = xgb_model
)
xgb.plot.importance(importance_matrix, top_n = 10,
                   main = "Feature Importance - XGBoost")
```
